Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-02-12 analysis/paper/paper.Rmd
added color to the plot

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022]. As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the simultaneous introduction of millet and rice during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022; @Weisskopf2015]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

@@

delta_meth_cor <-
  delta_meth_cor %>%
  slice(2:7) %>%
  dplyr::mutate(across(starts_with("13C"), as.numeric)) %>%
  dplyr::mutate(delta = `13C C18:0`-`13C C16:0`) %>%
  mutate(sample = case_when(
    C13 %in% c("Pot 1") ~ "CDG-062",
    C13 %in% c("Pot 2") ~ "SYG-3",
    C13 %in% c("Pot 3") ~ "SYG-TN13-E22-2#1",
    C13 %in% c("Pot 4") ~ "SYG-TN13-E22-1",
    C13 %in% c("Pot 5") ~ "SYG-TN13-E22-2#2",
    C13 %in% c("Pot 6") ~ "SYG-TN13-E23-3")) %>%
  mutate(sample = factor(sample, levels = c("SYG-TN13-E22-1",
                                            "SYG-TN13-E22-2#1",
                                            "SYG-TN13-E22-2#2",
                                            "SYG-TN13-E23-3",
                                            "SYG-3",
                                            "CDG-062")))

@@

# plot the carbon isotopes 
isotope_C16_C18 <-
  ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
       aes(`13C C16:0`,`13C C18:0`)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
  #ggrepel::geom_text_repel(aes(label = C13)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(delta*{}^13*"C 18:0 \u2030")) +
  scale_x_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  scale_y_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  #coord_fixed(ratio = 1) +
  geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
  geom_abline(intercept = -3.1, linetype = "dashed") +
  annotate("text", x = -21.5, y = -22, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -3.1 \u2030"))

big_delta_C16 <-
  ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
       aes(`13C C16:0`, delta)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
  scale_x_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  ylim(-2.5, 1.5) +
  geom_hline(yintercept = -0.7, linetype = "dashed") +
  annotate("text", x = -31.5, y = -0.9,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -32, y = 1.3,
           label = "C3-diet") +
  annotate("segment", x = -29, y = 1.3, xend = -24, yend = 1.3,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x = -21, y = 1.3,
           label = "C4-diet") +
  annotate("text", x = -22, y = -1.1,
           label = "Ruminant") +
  annotate("text", x = -21.5, y = -0.3,
           label = "Non-ruminant")

>>>>

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022]. As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the simultaneous introduction of millet and rice during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022; @Weisskopf2015]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

@@

delta_meth_cor <-
  delta_meth_cor %>%
  slice(2:7) %>%
  dplyr::mutate(across(starts_with("13C"), as.numeric)) %>%
  dplyr::mutate(delta = `13C C18:0`-`13C C16:0`) %>%
  mutate(sample = case_when(
    C13 %in% c("Pot 1") ~ "CDG-062",
    C13 %in% c("Pot 2") ~ "SYG-3",
    C13 %in% c("Pot 3") ~ "SYG-TN13-E22-2#1",
    C13 %in% c("Pot 4") ~ "SYG-TN13-E22-1",
    C13 %in% c("Pot 5") ~ "SYG-TN13-E22-2#2",
    C13 %in% c("Pot 6") ~ "SYG-TN13-E23-3")) %>%
  mutate(sample = factor(sample, levels = c("SYG-TN13-E22-1",
                                            "SYG-TN13-E22-2#1",
                                            "SYG-TN13-E22-2#2",
                                            "SYG-TN13-E23-3",
                                            "SYG-3",
                                            "CDG-062"))) %>% 
  mutate(period = ifelse(sample %in% c("SYG-TN13-E22-1", "SYG-3" ),
                         "Historical", "Neolithic"))

@@

# plot the carbon isotopes 
isotope_C16_C18 <-
  ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
       aes(`13C C16:0`,`13C C18:0`)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample, color = period)) +
  #ggrepel::geom_text_repel(aes(label = C13)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(delta*{}^13*"C 18:0 \u2030")) +
  scale_x_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  scale_y_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  #coord_fixed(ratio = 1) +
  geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
  geom_abline(intercept = -3.1, linetype = "dashed") +
  annotate("text", x = -21.5, y = -22, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -3.1 \u2030"))

big_delta_C16 <-
  ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
       aes(`13C C16:0`, delta)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample, color = period)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
  scale_x_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  ylim(-2.5, 1.5) +
  geom_hline(yintercept = -0.7, linetype = "dashed") +
  annotate("text", x = -31.5, y = -0.9,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -32, y = 1.3,
           label = "C3-diet") +
  annotate("segment", x = -29, y = 1.3, xend = -24, yend = 1.3,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x = -21, y = 1.3,
           label = "C4-diet") +
  annotate("text", x = -22, y = -1.1,
           label = "Ruminant") +
  annotate("text", x = -21.5, y = -0.3,
           label = "Non-ruminant")
